### 2020-06-11 manuscript.tex
edits 11.6.20

Cryptic species are present throughout the tree of life. They are especially prevalent in ferns, because of processes such hybridization, polyploidy, and reticulate evolution. In addition, the morphological simple body plan of ferns limits phenotypic variation and makes it difficult to detect cryptic species in ferns without molecular work. The model fern genus \textit{Ceratopteris} has long been suspected to harbor cryptic diversity, specifically in the highly polymorphic \textit{C. thalictroides}. Yet no studies have included samples from throughout the pan-tropical range of \textit{Ceratopteris} or utilized genomic sequencing, making it difficult to assess the full extent of cryptic variation within this genus. Here, we present the first multilocus phylogeny of the genus using reduced representation genomic sequencing (RADseq) and examine population structure, phylogenetic relationships, and ploidy level variation. We recover similar species relationships found in previous studies, find support for a named cryptic species as genetically distinct, and identify a novel putative species from within \textit{C. thalictroides} \textit{sensu latu} in Central and South America.

@@

Describing the formation of species is essential to understanding the patterns and processes that create biodiversity. Detecting differences among species becomes increasingly challenging when taxa are morphologically similar, but genetically distinct and reproductively isolated \autocite{Bickford2007, Masuyama1992, Paris1989}. Such cryptic species have often been historically described as one larger species, or species complex, due to morphological similarities \autocite{Paris1989}. However, advances in molecular methods have revealed that cryptic species can be monophyletic with disparate ecological niches and functions \autocite{Amato2007, Hebert2004, Sattler2007, Southgate2019}; in contrast, cryptic species complexes can be paraphyletic and separated by considerable evolutionary time, yet look similar and occupy comparable niches \autocite{Amor2014, Cunnington2005}.

Despite the recent shift in molecular approaches to investigate cryptic species, much is still unknown about their evolutionary history and ecosystem functions. Cryptic species complexes often have very broad distributions \autocite{Der2009, Knowlton1993, Nygren2014} and can occur in sympatry, allopatry, or parapatry \autocite{Bickford2007}, making adequate sampling challenging. In ferns, deciphering species boundaries can be even more difficult due to their ease of dispersal via spores, making gene flow possible across vast distances \autocite{Barrington1993, Tryon1970}. Compared to seed plants, ferns have a high prevalence of polyploidy and reticulate evolution \autocite{Barrington1989, Paris1989, Sigel2016, Otto2000, Otto2000}. Changes in ploidy can contribute to cryptic variation by altering niche space or offspring viability \autocite{Otto2007, Southgate2019, Masuyama2002}, yet have small effects on phenotype \autocite{Patel2019}.

The fern genus \textit{Ceratopteris} Brong. is a pan-tropical aquatic clade consisting of seven named species, three of which are cryptic and tetraploid \autocite{PPGI, LloydTax1974, Masuyama2010}. \textit{Ceratopteris} is perhaps best known for the model organism \textit{C. richardii}, which has been used as such since the late twentieth century \autocite{Banks1994, hickock1987, Hickok1995}. Sometimes called the ``\textit{Arabidopsis} of the fern world'' \autocite{Sessa2014}, \textit{C. richardii} is an ideal model system because of its fast life cycle and ease of cultivation \autocite{hickock1987}; in addition, it can be transformed with recombinant DNA \autocite{Muthukumar2013, Plackett2014}, has a reference ontogeny framework \autocite{Conway2019-cd}, and is currently the only homosporous fern to have a whole genome sequence \autocite{Marchant2019}. In addition to \textit{C. richardii}, other \textit{Ceratopteris} species have been studied in a lab environment \autocite{hickok1974, Hickok1977}, and have potential for further research. The species boundaries within the genus are blurry, however, and there is evidence that all species in the group (both diploid and tetraploid) hybridize to some extent \autocite{Adjie2007, hickok1974, Hickok1977, Hickok1979, LloydTax1974}. In such a well-utilized model genus, there is still a need to better understand species boundaries, evolutionary history, and occurrences of cryptic species. In particular, we need to understand the evolutionary dynamics of the \textit{Ceratopteris} genus as a whole in order to best utilize this model system for future work.

The first and only comprehensive monograph for \textit{Ceratopteris} was written by Lloyd in 1974, and employed a matrix of morphological traits to identify four species in the genus\nocite{LloydTax1974}. However, \textit{Ceratopteris} is a notorious group in terms of morphology: there are relatively few informative physical characters \autocite{LloydTax1974}; in addition, habitat and developmental stage can have a large effect on plant phenotype \autocite{Masuyama1992}. Most challenging for species delimitation is that rampant hybridization in the genus is known to further alter morphology and so make field identification difficult \autocite{hickok1974, LloydTax1974, Masuyama2010}. While Lloyd named four species, he noted that \textit{C. thalictroides}, the most widespread species in the genus, is ``highly polymorphic'' \autocite{LloydTax1974} and is likely a cryptic species complex. In response, Masuyama and colleagues conducted a series of studies on Asian plants under the name of \textit{C. thalictroides}, examining chloroplast DNA and cross-breeding \autocite{Masuyama2002}, cytological characteristics \autocite{Masuyama2005}, morphological traits \autocite{Masuyama1992, Masuyama2008}, and nuclear DNA \autocite{Adjie2007}. As a result, three cryptic species and two varieties were named from entities originally described as \textit{C. thalictroides} \autocite{Masuyama2010}. Molecular evidence suggests that all three cryptic species of \textit{C. thalictroides} have independent hybrid allopolyploid origins and are paraphyletic \autocite{Adjie2007}, the latter providing the most compelling evidence that they cannot be described under the same name. However, this study included only one nuclear marker; multilocus analyses are generally needed to substantiate paraphyly and polyploid origins \autocite{Eaton2013, Jorgensen2017}, and particularly in a case as challenging as \textit{C. thalictroides}. In addition, while cryptic species in \textit{Ceratopteris} have been investigated in the Old World, there is a similar problem of morphological variability in New World \textit{C. thalictroides} \autocite{Masuyama2010, LloydTax1974}, indicating a potential for more cryptic species to be discovered.

@@

To best analyze the cryptic diversity within \textit{Ceratopteris} we gathered samples from across its pan-tropical distribution. We utilized RADseq to generate a large genomic dataset, which was processed using the ipyrad pipeline \autocite{Eaton2020} for downstream analysis. Our three-part approach for assessing cryptic diversity takes advantage of the flexibility of RADseq data, and provides a unique window into the processes of speciation and diversification in a complex genus.

@@

Raw data were demultiplexed using stacks v. 2.4 \autocite{Catchen2011, Catchen2013} process\_radtags, allowing for a maximum of one mismatch per barcode. Demultiplexed FASTQ files were paired merged using ipyrad version 0.9.52 \autocite{Eaton2020}. Low quality bases, plus adapters and primers were removed from each read. Filtered reads were clustered at the default setting of 85\%, and we required a sequencing depth of 6 or greater per base and a minimum of 30 samples per locus to be included in the final assembly. The ipyrad pipeline defines a locus as a short sequence present across samples. From each locus, ipyrad identifies single nucleotide polymorphisms (SNPs); these SNPs are the variation used in downstream analyses. Since stacks and ipyrad assume diploidy, we processed all samples as such. It is well-known that some species of \textit{Ceratopteris} are non-diploid \autocite{Adjie2007, Masuyama2010}, and so we explore the ploidy of each sample more thoroughly later in our analyses.

Our initial sampling contained 89 specimens of \textit{Ceratopteris} and one sample of \textit{Acrostichum aureum}, plus six replicates. After checking for quality and similarity of replicates, they were combined with the corresponding samples. Of the 90 unique starting samples of \textit{Ceratopteris}, 26 either extracted very poorly or yielded few loci (less than 1000 loci per individual). These were removed from downstream analysis. We also removed 13 samples from Australia, an area we were able to sample on a very fine spatial scale: one individual per Australian population was included in the final dataset so we did not over represent Australian specimens in our analyses. Our final data set included 49 samples of \textit{Ceratopteris} and one sample of \textit{A. aureum}.  We did two separate runs of the ipyrad pipeline: one with \textit{A. aureum} for phylogenetic analysis, and one without for ploidy estimation and population structure analysis. Because \textit{A. aureum} is sister to \textit{Ceratopteris} and separated by over 30 million years \autocite{PPGI}, the total number of SNPs retrieved decreases when \textit{A. aureum} is included.

@@

Because species of \textit{Ceratopteris} are known to hybridize \autocite{hickok1974}, but also form reproductively isolated cryptic species \autocite{Masuyama2002}, we wanted to examine population structure of the genus across its distribution. The program {\small{STRUCTURE}} v. 2.3.4 \autocite{Pritchard2000} is designed to determine admixture (hybridization) between populations of one or several closely related taxa. The program assumes that each individual's genome is a mosaic from \textit{K} source populations and uses genotype assignments to infer population structure and admixture. In our analysis, we ran {\small{STRUCTURE}} for \textit{K} = 2 - 7 with 50 chains for each \textit{K}. We then used {\small{CLUMPAK}} \autocite{Kopelman2015} to process the {\small{STRUCTURE}} output and estimate the best \textit{K} values \autocite{Evanno2005, Pritchard2000}.

@@

To construct a phylogeny for the samples of \textit{Ceratopteris}, we used the program {\small{TETRAD}}, which is included in the ipyrad analysis toolkit and is based on the software \small{SVDQuartets} \autocite{Chifman2015}. {\small{TETRAD}} utilizes the theory of phylogenetic invariants to infer quartet trees from a SNP alignment. Species relationships are estimated for all quartet combinations of individuals, then all quartet trees are joined into a species tree with the software {\small{wQMC}} \autocite{Avni2015}. Quartet methods were designed to reduce computational time for tree-building with maximum likelihood methods \autocite{Ranwez2001}, making them particularly useful for SNP datasets such as these.

We ran {\small{TETRAD}} via Python 2.7.16 (www.python.org) for 56 samples of \textit{Ceratopteris}, and one sample of \textit{Acrostichum aureum} as the outgroup. Our input consisted of 57 individuals and 29849 SNPs; we ran 100 bootstrap iterations for the final consensus tree. We then plotted the final tree in R using the package phytools \autocite{Revell2012} and custom plotting functions.

% NOTE: add in mrbayes, raxml, and/or bucky here

@@

Due to the presence of numerous cryptic species and morphological variation across the genus, we hypothesize there may be cryptic cytotype variation and/or ploidy type (allo vs. auto) within species, especially the polymorphic \textit{C. thalictroides}. Since a majority our specimens were from herbarium specimens, we could not perform flow cytometry or chromosome squashes. We utilized the R v. 3.5.2 \autocite{R_352} package gbs2ploidy to detect ploidy variation across samples. This package was designed to detect ploidy levels in variable cytoptype populations (specifically quaking aspen, \textit{Populus tremuloides}) using low-coverage (2X) genotyping-by-sequencing (GBS) or RADseq data; it was also tested on simulated data, and can detect diploids, triploids, or tetraploids \autocite{Gompert2017}. This package estimates ploidy from allele ratios, which are calculated from genome-average heterozygosity and bi-allelic SNPs isolated from a variant calling format (VCF) file, which is included in the output from ipyrad.

To perform our analyses with gbs2ploidy, we first used the script vcf2hetAlleleDepth.py (github.com/carol-rowe666/vcf2hetAlleleDepth) to convert the VCF file produced by ipyrad to the format needed for gbs2ploidy. We then ran gbs2ploidy, using the function estprops to estimate ploidy for each individual. We plotted the output from gbs2ploidy in R, with the mean posterior probability for allelic ratios on the \textit{y} axis and the 1:1, 2:1, and 3:1 ratios on the \textit{x} axis (see Fig. \ref{g2p}); we also included error bars for the 95\% equal tail probability intervals (ETPIs). We assigned ploidy to each individual using the  highest posterior mean estimate for a certain allelic proportion. If the ETPIs overlapped, we considered it an ambiguous assignment. We required sequencing depth to be 6X or greater during our data processing, which suggests that the ploidy assignments are likely accurate.

@@

Our sampling included relatively few individuals of \textit{Ceratopteris richardii}, so we utilized the published genome from \textcite{Marchant2019} to perform an \textit{in silico} RADseq digest, creating a pseudo-indiviual to include in our analyses. We used the program ddRADseqTools \autocite{Mora-Marquez2017} to digest the genome with the same enzymes used in our \textit{in vivo} digest (PstI and BfaI). Then we performed a simulated ddRADseq run with paired-end reads, demultiplexing, and trimming using ddRADseqTools. This simulated individual was added to our other samples, processed via the ipyrad pipeline, and analyzed via STRUCTURE.

@@

We used the best \textit{K} method by \textcite{Evanno2005}, which determined a \textit{K} value of 4. The best \textit{K} estimate described in the STRUCTURE manual \autocite{Pritchard2000} determined a \textit{K} value of 5. We also visually examined \textit{K} = 2 - 7 and determined that \textit{K} = 5 was the most biologically meaningful because it separated each named species, and also showed some intra-specific variability (Fig. \ref{structure}). Increasing \textit{K} past 5 did not add any meaningful population clusters.

Named species of \textit{Ceratopteris} mostly clustered together at \textit{K} = 5 (Fig. \ref{structure}). \textit{Ceratopteris cornuta} and \textit{C. pteridoides} appeared as distinct populations. \textit{Ceratopteris thalictroides} \textit{sensu latu} split in to three groups at \textit{K} = 5. The first group consisted of several individuals identified as \textit{C. thalictroides} (on herbarium sheets or in the field) that clustered with \textit{C. gaudichaudii}, a cryptic species of \textit{C. thalictroides} \autocite{Masuyama2010}. The remaining individuals of \textit{C. thalictroides} grouped into two populations, one consisting of mostly Old World individuals, and the other as entirely New World individuals. These two populations were strongly differentiated by {\small{STRUCTURE}}, being separated at all \textit{K} values.
%Unfortunately, all samples of \textit{C. richardii} had very poor sequencing depth and quality, and their initial population assignments were variable; they were removed from the final STRUCTURE analysis.

@@

 The phylogeny resulting from our species quartet inference shows a similar pattern to the {\small{STRUCTURE}} output (Fig. \ref{phy}). Most individuals of \textit{Ceratopteris pteridoides} form a monophyletic group, with one individual sister to the remaining species. \textit{Ceratopteris cornuta} came out in two places in the phylogeny: sister to Old World \textit{C. thalictroides}, and in a grade outside the \textit{C. gaudichaudii}, Old World and New World \textit{C. thalictroides} clade. Our phylogeny grouped all individuals placed in the \textit{C. gaudichaudii} population by {\small{STRUCTURE}} in a monophyletic clade, sister to New World \textit{C. thalictroides}.

@@

\textit{Ceratopteris cornuta}, \textit{C. pteridoides}, and \textit{C. richardii} are known as a diploid (\textit{n} = 39) \autocite{Adjie2007, Hickok1977}; \textit{Ceratopteris thalictroides} and \textit{C. gaudichaudii} are known as tetraploid, with some cytotypic variation (\textit{n} = 77, 78) \autocite{Adjie2007, Masuyama2010}. Our ploidy analysis used allelic ratios to determine the ploidy of samples. Diploids would have a 1:1 allele ratio in accordance with the two parental genomes present. Triploids (derived via hybridization) would have a 2:1 allele ratio, with two sets of chromosomes from a unreduced spore, and another from a haploid spore. Tetraploids could either have a 1:1 or 3:1 allele ratio: the former would be representative of an autopolyploid derived from a genome double event, or an allopolyploid with two sets of homeologous chromosomes; the latter would be representation of an allopolyploid that has undergone genomic restructuring  (i.e. compensated aneuploidy, \cite{Sigel2016}).

Three of four samples identified as \textit{Ceratopteris cornuta} showed relatively equal allelic proportions of 1:1 and 3:1, indicating a possible tetraploid; the fourth sample was ambiguous. Some samples of the known diploid \textit{C. pteridoides} showed a similar pattern, with one putative diploid and three tetraploids. However, both species are known as a diploid so this may be a mis-assignment of ploidy by our analysis. \textit{Ceratopteris gaudichaudii} had a high 1:1 allele ratio, suggesting that it is an allotetraploid. \textit{Ceratopteris thalictroides} is also known to be a tetraploid \autocite{LloydTax1974, Masuyama2010}, and individuals in our analysis were estimated to have 1:1, 2:1, and 3:1 allele ratios. Despite this variation, there was a distinct pattern between Old World and New World individuals: Old World \textit{C. thalictroides} was most commonly allotetraploid (1:1 allele ratio), whereas New World \textit{C. thalictroides} was either auto- or allotetraploid (relatively equal 1:1 and 3:1 allele ratios).

%None of the individuals of \textit{C. richardii} could be assigned to a distinct ploidy, due to the poor data quality (see Fig. \ref{g2p}). We had a few other samples with very poor quality data (see Supplementary data 1), and their ploidy assignments were almost all ambiguous. 
Mis-assignment of ploidy could be due to coverage being at the lower end of the scale needed for true ploidy assignments, but it also could indicate hybridization leading to genomic complexity across species. Including chromosome counts and flow cytometry data in future analyses would help better understand the potential variation in polyploidy origins of \textit{C. thalictroides}.

@@

To the best of our knowledge, this is the first application of next-generation sequencing techniques to the genus \textit{Ceratopteris}. Using population genomic, phylogenetic, and ploidy inferences, we recover similar species relationships found in previous studies, find support for \textit{C. gaudichaudii} as a genetically distinct cryptic species, and identify a novel cryptic lineage from within \textit{C. thalictroides} \textit{sensu latu} in Central and South America.

@@

\textbf{4.1.1.\textit{Ceratopteris cornuta} and \textit{C. pteridoides}}

\textit{Ceratopteris cornuta} is found from western Africa to Northern Australia, and \textit{C. pteridoides} is endemic to the New World tropics \autocite{LloydTax1974}. These two species are the most morphologically conserved and distinct in the genus, and were recognized as distinct clades by our population structure analysis.

>>>>

Cryptic species are present throughout the tree of life. They are especially prevalent in ferns, because of processes such hybridization, polyploidy, and reticulate evolution. In addition, the simple morphology of ferns limits phenotypic variation and makes it difficult to detect cryptic species. The model fern genus \textit{Ceratopteris} has long been suspected to harbor cryptic diversity, specifically in the highly polymorphic \textit{C. thalictroides}. Yet no studies have included samples from throughout its pan-tropical range or utilized genomic sequencing, making it difficult to assess the full extent of cryptic variation within this genus. Here, we present the first multilocus phylogeny of the genus using reduced representation genomic sequencing (RADseq) and examine population structure, phylogenetic relationships, and ploidy level variation. We recover similar species relationships found in previous studies, find support for a named cryptic species as genetically distinct, and identify a novel genomic variation within two of the mostly broadly distributed species in the genus.

@@

Describing the formation of species is essential to understanding the patterns and processes that create biodiversity. Detecting differences among species becomes increasingly challenging when taxa are morphologically similar, but genetically distinct and reproductively isolated \autocite{Bickford2007, Masuyama1992, Paris1989}. Such cryptic species have often been historically described as one larger species, or species complex, due to morphological similarities \autocite{Paris1989}. However, advances in molecular methods have revealed that cryptic species can be monophyletic with disparate morphologies, ecological niches and functions \autocite{Amato2007, Hebert2004, Sattler2007, Southgate2019}. In addition, cryptic species complexes can also be paraphyletic and separated by considerable evolutionary time, yet look similar and occupy comparable niches \autocite{Amor2014, Cunnington2005}.

Despite the recent shift in molecular approaches to investigate cryptic species, much is still unknown about their evolutionary history and ecosystem functions. Cryptic species complexes often have very broad distributions \autocite{Der2009, Knowlton1993, Nygren2014} and can occur in sympatry, allopatry, or parapatry \autocite{Bickford2007}, making adequate sampling challenging. Studies have shown that cryptic species are prevalent across the Tree of Life (e.g. \cite{Bickford2007, Del_Carmen_Molina2011, Hebert2004, Nygren2014}), and ferns in particular are known to harbor many lineages with cryptic variation \autocite{Adjie2007, Paris1989, Yatabe2009}. However, deciphering species boundaries in ferns can be quite difficult due to their ease of dispersal via spores, making gene flow possible across vast distances \autocite{Barrington1993, Tryon1970}. Compared to seed plants, ferns have a high prevalence of polyploidy and reticulate evolution \autocite{Barrington1989, Paris1989, Sigel2016, Otto2000, Otto2000}. In addition, changes in ploidy can contribute to cryptic variation in ferns by altering niche space or offspring viability \autocite{Otto2007, Southgate2019, Masuyama2002}, yet have small effects on phenotype \autocite{Patel2019}. The widespread fern genus \textit{Ceratopteris} Brong. (Pteridaceae) is understood to have considerable cryptic variation, hybridization and polyploidy \autocite{Adjie2007, LloydTax1974, Masuyama2010}, making it an ideal system to investigate the origins and ecology of cryptic species.

The fern genus \textit{Ceratopteris} is a pan-tropical aquatic clade consisting of seven named species, three of which are cryptic and tetraploid \autocite{PPGI, LloydTax1974, Masuyama2010}. \textit{Ceratopteris} is perhaps best known for the model organism \textit{C. richardii}, which has been used as such since the late twentieth century \autocite{Banks1994, hickock1987, Hickok1995}. Sometimes called the ``\textit{Arabidopsis} of the fern world'' \autocite{Sessa2014}, \textit{C. richardii} is an ideal model system because of its fast life cycle and ease of cultivation \autocite{hickock1987}; in addition, it can be transformed with recombinant DNA \autocite{Muthukumar2013, Plackett2014}, has a reference ontogeny framework \autocite{Conway2019-cd}, and is currently the only homosporous fern to have a published genome sequence \autocite{Marchant2019}. In addition to \textit{C. richardii}, other \textit{Ceratopteris} species have been studied in a lab environment \autocite{hickok1974, Hickok1977}, and have potential for further research. The species boundaries within the genus are blurry, however, and there is evidence that all species in the group (both diploid and tetraploid) hybridize to some extent \autocite{Adjie2007, hickok1974, Hickok1977, Hickok1979, LloydTax1974}. In such a well-utilized model genus, there is still a need to better understand species boundaries, evolutionary history, and occurrences of cryptic species. In particular, we need to understand the evolutionary dynamics of the \textit{Ceratopteris} genus as a whole in order to best utilize this model system for future work.

The first and only comprehensive monograph for \textit{Ceratopteris} was written by Lloyd in 1974, and employed a matrix of morphological traits to identify four species in the genus\nocite{LloydTax1974}. However, \textit{Ceratopteris} is a notorious group in terms of morphology: there are relatively few informative physical characters \autocite{LloydTax1974}; in addition, habitat and developmental stage can have a large effect on plant phenotype \autocite{Masuyama1992}. Most challenging for species delimitation is that rampant hybridization in the genus is known to further alter morphology and so make field identification difficult \autocite{hickok1974, LloydTax1974, Masuyama2010}. While Lloyd named four species, he noted that \textit{C. thalictroides}, the most widespread species in the genus, is ``highly polymorphic'' \autocite{LloydTax1974} and is likely a cryptic species complex. In response, Masuyama and colleagues conducted a series of studies on Asian plants under the name of \textit{C. thalictroides}, examining chloroplast DNA and cross-breeding \autocite{Masuyama2002}, cytological characteristics \autocite{Masuyama2005}, morphological traits \autocite{Masuyama1992, Masuyama2008}, and nuclear DNA \autocite{Adjie2007}. As a result, three cryptic species and two varieties were named from entities originally described as \textit{C. thalictroides} \autocite{Masuyama2010}. Molecular evidence suggests that all three cryptic species of \textit{C. thalictroides} have independent hybrid allopolyploid origins and are paraphyletic \autocite{Adjie2007}, the latter providing the most compelling evidence that they cannot be described under the same name. However, that study included only one nuclear marker; multilocus analyses are generally needed to substantiate paraphyly and polyploid origins \autocite{Eaton2013, Jorgensen2017}, and particularly in a case as challenging as \textit{C. thalictroides}. In addition, while cryptic species in \textit{Ceratopteris} have been investigated in the Old World, there is a similar problem of morphological variability in New World \textit{C. thalictroides} \autocite{Masuyama2010, LloydTax1974}, indicating a potential for more cryptic species to be discovered.

@@

To best analyze the cryptic diversity within \textit{Ceratopteris} we gathered samples from across its pan-tropical distribution. We utilized RADseq to generate a large genomic dataset, which was processed using the \textit{ipyrad} pipeline \autocite{Eaton2020} for downstream analysis. Our three-part approach for assessing cryptic diversity takes advantage of the flexibility of RADseq data, and provides a unique window into the processes of speciation and diversification in a complex genus.

@@

Raw data were demultiplexed using stacks v. 2.4 \autocite{Catchen2011, Catchen2013} process\_radtags, allowing for a maximum of one mismatch per barcode. Demultiplexed FASTQ files were paired merged using \textit{ipyrad} version 0.9.52 \autocite{Eaton2020}. Low quality bases, plus adapters and primers were removed from each read. Filtered reads were clustered at the default setting of 85\%, and we required a sequencing depth of 6 or greater per base and a minimum of 30 samples per locus to be included in the final assembly. The \textit{ipyrad} pipeline defines a locus as a short sequence present across samples. From each locus, \textit{ipyrad} identifies single nucleotide polymorphisms (SNPs); these SNPs are the variation used in downstream analyses. Since stacks and \textit{ipyrad} assume diploidy, we processed all samples as such. It is well-known that some species of \textit{Ceratopteris} are non-diploid \autocite{Adjie2007, Masuyama2010}, and so we explore the ploidy of each sample more thoroughly later in our analyses.

Our initial sampling contained 89 specimens of \textit{Ceratopteris} and one sample of \textit{Acrostichum aureum}, plus six replicates. After checking for quality and similarity of replicates, they were combined with the corresponding samples. Of the 90 unique starting samples of \textit{Ceratopteris}, 26 either extracted very poorly or yielded few loci (less than 1000 loci per individual). These were removed from downstream analysis. We also removed 13 samples from Australia, an area we were able to sample on a very fine spatial scale: one individual per Australian population was included in the final dataset so we did not over represent Australian specimens in our analyses. Our final data set included 50 samples of \textit{Ceratopteris} and one sample of \textit{A. aureum}.  We did two separate runs of the \textit{ipyrad} pipeline: one with \textit{A. aureum} for phylogenetic analysis, and one without for ploidy estimation and population structure analysis. Because \textit{A. aureum} is sister to \textit{Ceratopteris} and separated by over 30 million years \autocite{PPGI}, the total number of SNPs retrieved for this individual was quite low.

@@

Because species of \textit{Ceratopteris} are known to hybridize \autocite{hickok1974}, but also form reproductively isolated cryptic species \autocite{Masuyama2002}, we wanted to examine population structure of the genus across its distribution. The program {\small{STRUCTURE}} v. 2.3.4 \autocite{Pritchard2000} is designed to determine admixture (hybridization) between populations of one or several closely related taxa. The program assumes that each individual's genome is a mosaic from \textit{K} source populations and uses genotype assignments from SNPs to infer population structure and admixture. We utilized a set of 8478 SNPs, created in \textit{pyrad} by selecting one SNP per loci. We ran {\small{STRUCTURE}} for \textit{K} = 2 - 6 with 50 chains for each \textit{K}. We then used {\small{CLUMPAK}} \autocite{Kopelman2015} to process the {\small{STRUCTURE}} output and estimate the best \textit{K} values \autocite{Evanno2005, Pritchard2000}.

@@

To construct a phylogeny for the samples of \textit{Ceratopteris}, we used the program {\small{TETRAD}}, which is included in the \textit{ipyrad} analysis toolkit and is based on the software \small{SVDQuartets} \autocite{Chifman2015}. {\small{TETRAD}} utilizes the theory of phylogenetic invariants to infer quartet trees from a SNP alignment. Species relationships are estimated for all quartet combinations of individuals, then all quartet trees are joined into a species tree with the software {\small{wQMC}} \autocite{Avni2015}. Quartet methods were designed to reduce computational time for tree-building with maximum likelihood methods \autocite{Ranwez2001}, making them particularly useful for SNP datasets such as these.

We ran {\small{TETRAD}} via Python 2.7.16 (www.python.org) for 50 samples of \textit{Ceratopteris}, and one sample of \textit{Acrostichum aureum} as the outgroup. Our input consisted of 51 individuals and 26891 SNPs; we ran 100 bootstrap iterations for the final consensus tree. As a comparison, we also ran {\small{RAxML}} v. 8.2.11 \autocite{Stamatakis2014} with 1000 bootstrap iterations. We then plotted the final trees in R using the package phytools \autocite{Revell2012} and custom plotting functions.

@@

Due to the presence of numerous cryptic species and morphological variation across the genus, we hypothesize there may be cryptic cytotype variation and/or ploidy type (allo vs. auto) within species, especially the polymorphic \textit{C. thalictroides}. Since a majority our specimens were from herbarium accessions, we could not perform flow cytometry or chromosome squashes. We utilized the R v. 3.5.2 \autocite{R_352} package gbs2ploidy to infer ploidy variation across samples. This package was designed to detect ploidy levels in variable cytoptype populations (specifically quaking aspen, \textit{Populus tremuloides}) using low-coverage (2X) genotyping-by-sequencing (GBS) or RADseq data; it was also tested on simulated data, and can detect diploids, triploids, or tetraploids \autocite{Gompert2017}. This package estimates ploidy from allele ratios, which are calculated from genome-average heterozygosity and bi-allelic SNPs isolated from a variant calling format (VCF) file, which is included in the output from \textit{ipyrad}.

To perform our analyses with gbs2ploidy, we first used the script vcf2hetAlleleDepth.py (github.com/carol-rowe666/vcf2hetAlleleDepth) to convert the VCF file produced by \textit{ipyrad} to the format needed for gbs2ploidy. We then ran gbs2ploidy, using the function estprops to estimate ploidy for each individual. We plotted the output from gbs2ploidy in R, with the mean posterior probability for allelic ratios on the \textit{y} axis and the 1:1, 2:1, and 3:1 ratios on the \textit{x} axis (see Fig. \ref{g2p}); we also included error bars for the 95\% equal tail probability intervals (ETPIs). We assigned ploidy to each individual using the  highest posterior mean estimate for a certain allelic proportion. If the ETPIs overlapped, we considered it an ambiguous assignment. We required sequencing depth to be 6X or greater during our data processing, which suggests that the ploidy assignments are likely accurate.

@@

Our sampling included relatively few individuals of \textit{Ceratopteris richardii}, so we utilized the published genome from \textcite{Marchant2019} to perform an \textit{in silico} RADseq digest, creating a pseudo-individual to include in our analyses. We used the program ddRADseqTools \autocite{Mora-Marquez2017} to digest the genome with the same enzymes used in our \textit{in vivo} digest (PstI and BfaI). Then we performed a simulated ddRADseq run with paired-end reads, demultiplexing, and trimming using ddRADseqTools. This simulated individual was added to our other samples, processed via the \textit{ipyrad} pipeline, and analyzed via STRUCTURE.

@@

To determin the best \textit{K} value across {\small{STRUCTURE}} we combined several metrics. We used the best \textit{K} method by \textcite{Evanno2005}, which determined a \textit{K} value of 4. The best \textit{K} estimate described in the {\small{STRUCTURE}} manual \autocite{Pritchard2000} determined a \textit{K} value of 5. We also visually examined \textit{K} = 2 - 7 and determined that \textit{K} = 5 was the most biologically meaningful because it separated each named species, and also showed some intra-specific variability (Fig. \ref{structure}). Increasing \textit{K} past 5 did not add any meaningful population clusters.

Named species of \textit{Ceratopteris} mostly clustered together at \textit{K} = 5 (Fig. \ref{structure}). \textit{Ceratopteris cornuta} and \textit{C. pteridoides} appeared as distinct populations, although there was some minor introgression from \textit{C. pteridoides} into the \textit{C. cornuta} population. \textit{Ceratopteris thalictroides} \textit{sensu latu} split in to three groups at \textit{K} = 5. The first group consisted of several individuals identified as \textit{C. thalictroides} (on herbarium sheets or in the field) that clustered with \textit{C. gaudichaudii}, a cryptic species of \textit{C. thalictroides} \autocite{Masuyama2010}. The remaining individuals of \textit{C. thalictroides} grouped into two populations, one consisting of mostly Old World individuals, and the other as entirely New World individuals. These two populations were strongly differentiated by {\small{STRUCTURE}}, being separated at all \textit{K} values.

@@

 The phylogeny resulting from our species quartet inference shows a similar pattern to the {\small{STRUCTURE}} output (Fig. \ref{phy}). Individuals of \textit{Ceratopteris pteridoides} form a monophyletic group, sister to the rest of the genus.z \textit{Ceratopteris cornuta} came out in two places in the phylogeny: sister to Old World \textit{C. thalictroides}, and in a grade outside the \textit{C. gaudichaudii}, Old World and New World \textit{C. thalictroides} clade. Our phylogeny grouped all individuals placed in the \textit{C. gaudichaudii} population by {\small{STRUCTURE}} in a monophyletic clade, sister to New World \textit{C. thalictroides}.

 The {\small{RAxML}} tree shows a very similar structure to the {\small{TETRAD}} tree. The only major difference was that instead of New World \textit{C. thalictroides} and \textit{C. gaudichaudii} forming a monophyletic group, the former comes out as sister to all other species except \textit{C. pteridoides}. Graphical comparison of the two trees is included in Supplementary data 2.

@@

\textit{Ceratopteris cornuta}, \textit{C. pteridoides}, and \textit{C. richardii} are known as a diploid (\textit{n} = 39) \autocite{Adjie2007, Hickok1977}; \textit{Ceratopteris thalictroides} and \textit{C. gaudichaudii} are known as tetraploid, with some cytotypic variation (\textit{n} = 77, 78) \autocite{Adjie2007, Masuyama2010}. Our ploidy analysis used allelic ratios to estimate the ploidy of samples. Diploids would have a 1:1 allele ratio in accordance with the two parental genomes present. Triploids (derived via hybridization) would have a 2:1 allele ratio, with two sets of chromosomes from a unreduced spore, and another from a haploid spore. Tetraploids could either have a 1:1 or 3:1 allele ratio: the former would be representative of an autopolyploid derived from a genome double event, or an allopolyploid with two sets of homeologous chromosomes; the latter would be representation of an allopolyploid that has undergone genomic restructuring  (i.e. compensated aneuploidy, \cite{Sigel2016}).

Three of four samples identified as \textit{Ceratopteris cornuta} showed relatively equal allelic proportions of 1:1 and 3:1, indicating a possible tetraploid; the fourth sample was ambiguous. Some samples of the known diploid \textit{C. pteridoides} showed a similar pattern, with one putative diploid and three tetraploids. However, both species are known as a diploid so this may be a mis-assignment of ploidy by our analysis, or detecting of some interesting genomic structure. \textit{Ceratopteris gaudichaudii} had a high 1:1 allele ratio, suggesting that it is an allotetraploid. \textit{Ceratopteris thalictroides} is also known to be a tetraploid \autocite{LloydTax1974, Masuyama2010}, and individuals in our analysis were estimated to have 1:1, 2:1, and 3:1 allele ratios. Despite this variation, there was a distinct pattern between Old World and New World individuals: Old World \textit{C. thalictroides} was most commonly had a 1:1 allele ratio (autopolyploid), whereas New World \textit{C. thalictroides} had  3:1 allele ratios (allopolyploid).

Mis-assignment of ploidy could be due to coverage being at the lower end of the scale needed for true ploidy assignments, but it also could indicate hybridization leading to genomic complexity across species. Including chromosome counts and flow cytometry data in future analyses would help better understand the potential variation in polyploidy origins of \textit{C. thalictroides}.

@@

To the best of our knowledge, this is the first application of next-generation sequencing techniques to the genus \textit{Ceratopteris}. Using population genomic, phylogenetic, and ploidy inferences, we recover similar species relationships found in previous studies, find support for \textit{C. gaudichaudii} as a genetically distinct cryptic species, identify a novel cryptic lineage from within \textit{C. thalictroides} \textit{sensu latu} in Central and South America, and show strong genetic variation across the broad range of \textit{C. coruta}.

@@

\textbf{4.1.1. Variation within \textit{Ceratopteris cornuta}}
